FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories

### 2024-11-19 manuscript.Rmd
update 2024-11-19 16:28:15.604972

In a progressive research program [@lakatosHistoryScienceIts1971],
the cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
There are indications that contemporary psychology falls short of this idealized model, however.
Firstly, because deductive research is over-represented in the literature. 
According to one estimate, 89.6% of published studies tests hypotheses,
which suggests that the literature is predominantly comprised of deductive research  [@kuhbergerPublicationBiasPsychology2014].
Closer examination reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies reference theory at all [@mcphetresDecadeTheoryReflected2021].
This invites the question where the other hypotheses come from,
and to what extent testing an ad-hoc hypothesis not grounded in theory can advance our principled understanding of the studied phenomena.
These statistics suggest that theory has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely derived from theory,
and test results rarely contribute to the improvement of existing theories.
Consequently, theories either persist unchanged, immune to refutation,
or are forgotten [REF Meehl].

@@

One major obstacle to cumulative knowledge acquisition is the difficulty in finding relevant theories.
The primary unit of dissemination and search in psychology is often the academic paper.
A paper may contain multiple resources - including materials, data, code, and theory - but there is no unified search engine for theory,
or standardized metadata to signal the presence of theory within a work,
nor even a standardized keyword (model, framework, etc are often used interchangeably with theory).
Making theories more findable would allow them to be systematically indexed, tagged, and made searchable.
This would allow researchers to easily identify relevant theories to inform their hypotheses,
grounding their work in established theoretical foundations.
Making theories findable also increases the impact and reuse potential of theories across scientific contexts.
For example, it becomes possible to study theories (the same way libraries and search engines have enabled scholars to study the literature via systematic reviews);
theories might have transdisciplinary impact;
and the structure of theories from one discipline might inspire analogical modeling in other fields, the way - for example, predator-prey models have inspired theories of XXX and the Eysenck model of atomic magnetism has inspired a network theory of depression.

One effort that improves theories' findability is post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website [REF GRAY].
Similarly, Borsboom and colleagues seek to establish a dictionary of psychological "phenomena" (which are not strictly theories, but are patterns reliably evidenced by data that theory should seek to explain) [REF BORSBOOM].
Another method that could ameliorate theory Findability is modular publishing.
In this approach, distinct assets are individually published as citable academic output.
Such output can be linked; for example, a FAIR theory could have an associated paper.

With regard to Accessibility, publishing behind paywalls certainly has a negative impact.
Open Access publishing increases the accessibility of all academic output, including theory.
Another factor curtailing the accessibility of theory is ambiguity;
a lack of theory formalization introduces a dependency on the original author for clarification.
The discourse on "Great Man Theorizing" touches upon the problems this introduces [REF Guest et al moral theory].
For example, dependency on interpretation by the author creates a potential for gatekeeping - the author could insist that work requires their involvement, which violates checks and balances of scientific research.
Moreover, if a theory is refuted, its author could claim that the author of the refuting paper did not interpret the theory correctly.
To some extent, this relates to the problem of translation [REF Duhem]:
it is not possible to entirely formalize an idea to enable unambiguous interpretation.
Nonetheless, taking care to formalize a theory to the maximum extent possible advances Accessibility.

@@

@@

2. Accessibility: Promoting Open and Transparent Theoretical Development
Accessibility involves ensuring that theories are openly available to all researchers, not locked behind paywalls or isolated in specific journals. By making theories FAIR, they would be accessible under clear, standardized, and open conditions, using platforms that allow both humans and machines to retrieve them effortlessly.

How it helps: Openly accessible theories would allow for greater scrutiny and replication by the research community, fostering transparent scholarly communication. This ensures that researchers worldwide can contribute to, refine, and build upon existing theories, thus accelerating the pace of scientific discovery and enhancing the collective understanding of a given phenomenon.
3. Interoperability: Facilitating Integration Across Fields
In many scientific disciplines, researchers often operate in isolated silos, using different terminologies, methodologies, and conceptual frameworks. The interoperability aspect of FAIR focuses on the ability of data (or in this case, theories) to be integrated with other systems using common standards, formats, and vocabularies.

>>>>

In a progressive research program [@lakatosHistoryScienceIts1971],
the cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
There are indications that contemporary psychology falls short of this idealized model, however.
Firstly, because deductive research is over-represented in the literature: 
According to one estimate, 89.6% of paper published in psychology tests hypotheses [@kuhbergerPublicationBiasPsychology2014].
Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies referenced any theory, and theory was often not references in connection to the hypothesis [@mcphetresDecadeTheoryReflected2021].
This invites the question where hypotheses come from in the remaining 85% of deductive studies,
and to what extent testing ad-hoc hypotheses not grounded in theory can advance our principled understanding of psychological phenomena.
These statistics suggest that theory has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely derived from theory,
and test results rarely contribute to the improvement of existing theories.
The paradoxical role of theory in psychology may help explain the observation that "theories either persist unchanged, immune to refutation,
or are forgotten" [check quote REF Meehl].

@@

One major obstacle to cumulative knowledge acquisition is the difficulty in finding relevant theories.
The primary unit of dissemination and search in psychology is often the academic paper.
A paper may contain multiple resources - including materials, data, code, and theory - but there is no unified search engine for theory,
nor standardized metadata or even a keyword to signal the presence of theory within a paper.
Making theories more findable would allow them to be systematically indexed, tagged, and made searchable.
This would allow researchers to easily identify relevant theories to inform their hypotheses,
grounding their work in established theoretical foundations.
Making theories findable also increases the impact and reuse potential of theories across scientific contexts.
For example, it becomes possible to conduct meta-research on theories, 
in the same way libraries and search engines have enabled scholars to study the literature via systematic reviews.
Moreover, theories might have transdisciplinary impact, either through direct application or through analogical modeling.
In analog modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
For example, predator-prey models have inspired theories of XXX, and the Eysenck model of atomic magnetism has inspired a network theory of depression.

One effort that improves theories' findability is post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website [REF GRAY].
Similarly, Borsboom and colleagues seek to establish a database of psychological theories [REF BORSBOOM].
Post-hoc curation does not address the root cause of the lack of Findability, however.
This would be achieved by modular publishing of theories as individually citable academic assets, with adequate metadata that is indexed in standardized repositories,
similar to the current practice of publishing empirical data in standardized repositories (e.g., DataVerse).
As with empirical data, these theories could still be connected to a specific paper which might serve as documentation and the canonical reference for the resource.

### Accessibility

Transparent scholarly communication about theory requires that theories are accessible to all researchers and other stakeholders.
Publishing theories behind paywalls impedes this process.
Open Access publishing increases the accessibility of all academic output, including theory.
If theories were more accessible, researchers could reuse and refine them, thus accelerating cumulative knowledge acquisition.
Making theories accessible also allows other stakeholders (e.g., practitioners, policy makers, advocates) to inform themselves of the current scientific understanding of specific phenomena:
More so than single studies, theories offer a top-down, big picture representation of the phenomena studied in a field.

Beyond open access publishing, making theories accessible also requires communicating them clearly and explicitly.
This relates to a property of good theories that has been dubbed "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024].
But contemporary psychological theory is often difficult to understand due to its ambiguity [REF Frankenhuis].
Ambiguity reduces the portability of theory by introducing a dependency on the original author for interpretation and clarification.
This dependency can be an obstacle, for example, when a theory is insufficiently clear to derive hypotheses from it without first asking the original author what they meant - but it can also render theories immune to refutation.
When confronted with falsifying evidence, the original author can always claim that the theory was misconstrued or misrepresented, thus making it a moving target.
Dependency on the author relates to the discourse on "Great Man Theorizing" [REF Guest et al moral theory].
For example, because dependency on the author creates a potential for gatekeeping - the author could insist that work requires their involvement, which violates checks and balances of scientific research.
Moreover, if a theory is refuted, its author could claim that the author of the refuting paper did not interpret the theory correctly.
Dependency on the author also relates to the problem of translation [REF Duhem]:
it is not possible to entirely formalize an idea to the point that it becomes unambiguously interpretable.
Despite this theoretical upper bound on formalization,
efforts to formalize a theory will advance its Accessibility.

@@

### Interoperability

@@

3. Interoperability: Facilitating Integration Across Fields
In many scientific disciplines, researchers often operate in isolated silos, using different terminologies, methodologies, and conceptual frameworks. The interoperability aspect of FAIR focuses on the ability of data (or in this case, theories) to be integrated with other systems using common standards, formats, and vocabularies.

### 2024-11-19 manuscript.tex
update 2024-11-19 16:28:15.604972

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 2657}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

This is a preprint paper, generated from Git Commit \# d1a82fc7.

@@

The ``replication crisis'' has prompted extensive reforms in social science (Lavelle, 2021; Scheel, 2022).
Concern that undisclosed flexibility in analyses was to blame for the abundance of false-positive findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
But have these reforms met their goal?
Recent reviews show that most preregistered hypothesis tests are not supported (Scheel, Schijen, \& Lakens, 2021).
One plausible explanation is that the replication crisis is symptomatic of an underlying,
and more fundamental, ``theory crisis''.
If psychological theories are insufficiently precise to derive testable hypotheses,
or sufficiently vague to explain contradictory findings,
then preregistration and replication will only serve to highlight those shortcomings.

Scholars have raised concerns about the state of theory in social science for nearly 50 years (Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
One contributing factor is a lack of theory formalization:
social scientific theories often lack the precision and clarity of theories in the physical sciences (Szollosi \& Donkin, 2021).
A second factor, which received less attention, is the a lack of transparent and democratic scholarly communication about psychological theory.
The present paper seeks to advance transparent communication about theory by applying open science principles to psychological theory for the first time,
and introducing the concept of \emph{FAIR Theory}.

@@

In a progressive research program (Lakatos, 1971),
the cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
There are clear indications that contemporary psychology falls short of this idealized model, however.
Firstly, because deductive research is over-represented in the literature.
According to one estimate, 89.6\% of published studies tests hypotheses,
which suggests that the literature is predominantly comprised of deductive research (Kühberger, Fritz, \& Scherndl, 2014).
Closer examination reveals, however, that the link between theory and hypothesis is often tenuous (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
Only 15\% of deductive studies reference theory at all (McPhetres et al., 2021).
This raises the question where the other hypotheses come from,
and what consequence their refutation would have for theory.
These statistics suggest that theory has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely derived from theory,
and test results rarely contribute to the improvement of existing theories.
Consequently, theories either persist unchanged for decades, or are forgotten {[}REF Meehl{]}.

Scientific reform initiated by the open science movement has predominanty focused on improving deductive methods, overlooking the shortcomings of theory.
The present paper applies, for the first time, open science principles to theory.

@@

\subsection{Is Current Psychological Theory FAIR?}\label{is-current-psychological-theory-fair}

While a comprehensive analysis of the present state of FAIR Theory in psychology is beyond the scope of the present paper,
we provide a narrative review of examples and counterexamples for each criterion.

One factor that contributes to poor findability of psychological theory is that the primary unit of dissemination and search in psychology is the academic paper.
A paper may contain multiple resources - including materials, data, code, and theory - but there is no unified search engine for theory,
or even an agreed-upon keyword (model, framework, etc are often used interchangeably with theory).
Modular publishing could ameliorate the Findability of psychological theory.
In this approach, distinct resources are individually published as citable academic output.
Such output can still be linked; for example, a FAIR theory could have an accompanying paper.
Another effort to improve theories' findability is post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website.
Similarly, Borsboom and colleagues seek to establish a dictionary of psychological ``phenomena'' (which are not strictly theories, but are patterns reliably evidenced by data that theory should seek to explain).

With regard to Accessibility, publishing behind paywalls certainly has a negative impact.
Open Access publishing increases the accessibility of all academic output, including theory.
Another factor curtailing the accessibility of theory is ambiguity;
a lack of theory formalization introduces a dependency on the original author for clarification.
The discourse on ``Great Man Theorizing'' touches upon the problems this introduces {[}REF Guest et al moral theory{]}.
For example, dependency on interpretation by the author creates a potential for gatekeeping - the author could insist that work requires their involvement, which violates checks and balances of scientific research.
Moreover, if a theory is refuted, its author could claim that the author of the refuting paper did not interpret the theory correctly.
To some extent, this relates to the problem of translation {[}REF Duhem{]}:
it is not possible to entirely formalize an idea to enable unambiguous interpretation.
Nonetheless, taking care to formalize a theory to the maximum extent possible advances Accessibility.

There appears to be a norm \emph{against} the reuse of theory,
as evident from the quip that
\emph{``{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's''} (Mischel, 2008).
Moreover, a legal basis for theory reuse is often absent.
Questions that often come up in our workshops on FAIR Theory are ``who owns theory'',
and ``who determines how a theory may be changed''?
Such questions require a clear answer, even if that answer may vary across theories.
Licensing theories for reuse provides clarity about how a theory can be reused.
A crucial legal consideration is that,
while specific (FAIR) representations of theory might be protected by copyright,
the underlying ideas are not.
This is another argument in favor of publishing specific instantiations of theory along with licenses detailing how they may be reused.

@@

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 3547}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

This is a preprint paper, generated from Git Commit \# 4d935da1.

@@

\subsection{The Need for FAIR Theory}\label{the-need-for-fair-theory}

The so-called ``replication crisis'' has prompted extensive reforms in social science (Lavelle, 2021; Scheel, 2022).
Concern that undisclosed flexibility in analyses was to blame for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
These various practices ensure transparent and repeated testing of hypotheses.
However, recent reviews show that most preregistered hypothesis tests are not supported (Scheel, Schijen, \& Lakens, 2021).
Increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
psychological theories rarely produce hypotheses that are found to be true,
and are often sufficiently vague to explain contradictory findings.

Scholars have raised concerns about the state of theory in social science for nearly 50 years (Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
One concern is that social scientific theories often lack the precision and clarity of theories in the physical sciences, otherwise known as \emph{formalization} (Szollosi \& Donkin, 2021).
A second concern, which received less attention, is the lack of transparent and democratic scholarly communication about psychological theory.
The present paper focuses on the second concern,
advancing transparent communication about theory by applying open science principles to psychological theory for the first time
and introducing the concept of \emph{FAIR Theory}.

@@

In a progressive research program (Lakatos, 1971),
the cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
There are indications that contemporary psychology falls short of this idealized model, however.
Firstly, because deductive research is over-represented in the literature.
According to one estimate, 89.6\% of published studies tests hypotheses,
which suggests that the literature is predominantly comprised of deductive research (Kühberger, Fritz, \& Scherndl, 2014).
Closer examination reveals, however, that the link between theory and hypothesis is often tenuous (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
Only 15\% of deductive studies reference theory at all (McPhetres et al., 2021).
This invites the question where the other hypotheses come from,
and to what extent testing an ad-hoc hypothesis not grounded in theory can advance our principled understanding of the studied phenomena.
These statistics suggest that theory has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely derived from theory,
and test results rarely contribute to the improvement of existing theories.
Consequently, theories either persist unchanged, immune to refutation,
or are forgotten {[}REF Meehl{]}.

\subsection{The Solution}\label{the-solution}

Applying the FAIR principles to scientific theories facilitates transparent scholarly communication and accelerates cumulative knowledge acquisition for several reasons.

\subsubsection{Findability}\label{findability}

One major obstacle to cumulative knowledge acquisition is the difficulty in finding relevant theories.
The primary unit of dissemination and search in psychology is often the academic paper.
A paper may contain multiple resources - including materials, data, code, and theory - but there is no unified search engine for theory,
or standardized metadata to signal the presence of theory within a work,
nor even a standardized keyword (model, framework, etc are often used interchangeably with theory).
Making theories more findable would allow them to be systematically indexed, tagged, and made searchable.
This would allow researchers to easily identify relevant theories to inform their hypotheses,
grounding their work in established theoretical foundations.
Making theories findable also increases the impact and reuse potential of theories across scientific contexts.
For example, it becomes possible to study theories (the same way libraries and search engines have enabled scholars to study the literature via systematic reviews);
theories might have transdisciplinary impact;
and the structure of theories from one discipline might inspire analogical modeling in other fields, the way - for example, predator-prey models have inspired theories of XXX and the Eysenck model of atomic magnetism has inspired a network theory of depression.

One effort that improves theories' findability is post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website {[}REF GRAY{]}.
Similarly, Borsboom and colleagues seek to establish a dictionary of psychological ``phenomena'' (which are not strictly theories, but are patterns reliably evidenced by data that theory should seek to explain) {[}REF BORSBOOM{]}.
Another method that could ameliorate theory Findability is modular publishing.
In this approach, distinct resources (data, code, instruments, theory) are individually published as citable academic assets.
Such output can be linked to a specific paper.

\subsubsection{Accessibility}\label{accessibility}

Advancing transparent scholarly communication about theory requires that theories are accessible to all researchers and other stakeholders.
Publishing behind paywalls clearly creates obstacles.
Open Access publishing increases the accessibility of all academic output, including theory.
If theories were more accessible, researchers could reuse, refine, and relate their work to those theories, thus accelerating cumulative knowledge acquisition.
Making theories more accessible also allows other stakeholders (e.g., practitioners, policy makers, advocates) to inform themselves of the current scientific understanding of specific phenomena:
More so than single studies, theories offer a top-down, big picture representation of the phenomena studied in a field.

Beyond open access publishing, making theories accessible also requires communicating them clearly and explicitly.
This property has been dubbed ``discursive survival {[}\ldots{]}, the ability to be understood'' (\textbf{guestWhatMakesGood2024?}).
Yet contemporary psychological theory is often ambiguous {[}REF Frankenhuis{]}.
Ambiguity introduces a dependency on the original author for clarification,
because the author can always claim that his ideas were misconstrued or misrepresented.
This discourse on ``Great Man Theorizing'' touches upon the problems this introduces {[}REF Guest et al moral theory{]}.
For example, dependency on interpretation by the author creates a potential for gatekeeping - the author could insist that work requires their involvement, which violates checks and balances of scientific research.
Moreover, if a theory is refuted, its author could claim that the author of the refuting paper did not interpret the theory correctly.
To some extent, this relates to the problem of translation {[}REF Duhem{]}:
it is not possible to entirely formalize an idea to enable unambiguous interpretation.
Nonetheless, taking care to formalize a theory to the maximum extent possible advances Accessibility.

The ultimate goal of the FAIR principles is to facilitate the reuse digital research artefacts.
While theory is ostensibly held in high regard in the social sciences,
there also appears to be a norm \emph{against} theory reuse.
This is evident from the quip that
\emph{``{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's''} (Mischel, 2008).
In our workshops on FAIR Theory, we often hear questions such as ``who owns theory'',
and ``who determines how a theory may be reused or changed''?
Licensing theories for reuse provides an unambiguous answer to such questions.

One important consideration is that there may be legal limits to the rights authors can claim over theory.
An important concept in copyright law is the \emph{idea-expression dichotomy} {[}REF{]}.

\begin{enumerate}
\def\labelenumi{\alph{enumi}.}
\setcounter{enumi}{14}
\tightlist
\item
  Ideas, facts, and concepts are not protected by Although they are not protectable by copyright, the expression of those ideas, facts, and concepts are protectable, such as in a description, explanation, or illustration or as a database of facts. Separating an idea from the expression or manifestation of that idea is known in copyright law as the
  Copyright explicitly does not
  \emph{``extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery''},
  but it may extend to creative works expressing that idea (e.g., writing, visual illustrations).
  Copyright pertaining to recipes offer an interesting perspective:
  Whereas the recipes themselves are not protected, \emph{``substantial literary expression in the form of an explanation or directions''} can be protected by copyright.
  Extrapolating to FAIR Theory, it would seem that vague, ambiguous verbal explanations of theories are more likely to qualify for copyright protection than formal theories.
  If copyright does not cover (formal) theory,
  then it would be pointless to adopt a license that assumes copyright protection.
  Furtermore, even if copyright would apply, research is covered under ``fair use'' excemptions to copyright law, so copyright would impose few restrictions to theory reuse in scholarly communication.
  In sum, the CC0 (no rights reserved) license seems most appropriate for FAIR Theory;
  it explicitly waives all rights and encourages reuse.
  Simultaneously, to comply with the interoperability principle, the principle of comprehensive citation {[}REF TOP guidelines{]},
  and the spirit of good scholarship,
  it is essential that scholars do comprehensively cite theory,
  including prior work that new theories are based on.
\end{enumerate}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  Interoperability: Facilitating Integration Across Fields
  In many scientific disciplines, researchers often operate in isolated silos, using different terminologies, methodologies, and conceptual frameworks. The interoperability aspect of FAIR focuses on the ability of data (or in this case, theories) to be integrated with other systems using common standards, formats, and vocabularies.
\end{enumerate}

How it helps: If theories are made interoperable, they can be used in multi-disciplinary research and can better interface with diverse fields. This reduces duplication of effort and enables researchers from different areas to combine and compare theories, bridging gaps between disciplines and fostering innovative, cross-disciplinary approaches. Researchers can easily synthesize theoretical perspectives, generating new hypotheses that are more comprehensive and grounded in a wider range of evidence.
4. Reusability: Enabling Continuous Improvement of Theories
For knowledge to accumulate meaningfully, theories need to be adaptable and subject to continuous revision as new data emerge. The reusability principle calls for thorough documentation and community standards to ensure that data (or theories) can be applied in new contexts.

How it helps: Reusable theories would provide clear metadata, version histories, and supporting evidence, allowing other researchers to test, refine, or extend them as new empirical findings emerge. This improves the feedback loop between theory and experimentation, ensuring that test results inform future theoretical development. It encourages the use of theories not just for their original purpose but in new and evolving contexts, thus ensuring that the theory remains dynamic and continuously refined over time.
Conclusion: Making Theories FAIR Accelerates Scientific Progress
Adopting the FAIR principles for theories can address key challenges in the current research landscape, where theories often remain isolated and underutilized. By making theories findable, accessible, interoperable, and reusable, researchers can ensure that their work is grounded in a shared, transparent, and cumulative body of knowledge. This approach enhances scholarly communication, allowing for greater scrutiny, replication, and collaboration across disciplines, ultimately leading to faster, more reliable, and more impactful scientific progress.

@@

@@

\section{Discussion}\label{discussion}

\subsection{Future Directions}\label{future-directions}

One remaining issue that intersects with FAIR Theory is the measurement and operationalization of psychological constructs.
Aside from the aforementioned ``theory crisis'', there has been talk of a ``measurement crisis'':
it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties {[}REF{]}.
Additionally, the ``jingle-jangle'' fallacy is prevalent in the social sciences:
the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
FAIR Theory can help address the measurement crisis:
since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of
